REVISITING SEMI-CONTINUOUS HIDDEN MARKOV MODELS

### 2015-09-03 misc/papers/icassp12_tied/icassp12.tex
changes to cudamatrix code to upgrade to the CuBLAS v2 API

\begin{abstract}
In the past decade, semi-continuous hidden Markov models (SC-HMMs) have not 
attracted much attention in the speech recognition community. Growing amounts 
of training data and increasing sophistication of model estimation led to the 
impression that continuous HMMs are the best choice of acoustic model.
%
However, recent work on recognition of under-resourced languages faces the same
old problem of estimating a large number of parameters from limited amounts 
of transcribed speech.
This has led to a renewed interest in methods of reducing the number of parameters 
while maintaining or extending the modeling capabilities of continuous models.
%
In this work, we compare classic and multiple-codebook semi-continuous models
using diagonal and full covariance matrices with continuous HMMs and subspace Gaussian 
mixture models.  
Experiments using on the RM and WSJ corpora show that while a classical semi-continous
system does not perform as well as a continuous one, multiple-codebook semi-continuous
systems can perform better, particular when using full-covariance Gaussians.
\end{abstract}

@@

We were motivated by a number of factors to look again at semi-continuous models.
One is that we are interested in applications where there is a small amount
of training data (or a small amount of in-domain training data), and since
in semi-continuous models the amount of state-specific parameters is relatively
small, we felt that they might have an advantage there.  We were also interested
in improving the open-source speech recognition toolkit {\sc Kaldi}~\cite{povey2011tks},
and felt that since a multiple-codebook form of semi-continous models is, to
our knowledge, currently being used in a state-of-the-art system~\cite{prasad2004t2b},
we should implement this type of model.  We also wanted to experiment with
the style of semi-continous system described 
in \cite{schukattalamazzini1994srf,schukattalamazzini1995as}, particularly the smoothing
techniques, and see how these compare against a more conventional system.
Lastly, we felt that it would be interesting to compare semi-continous
models with Subspace Gaussian Mixture Models (SGMMs)~\cite{povey2011sgm}, since
there are obvious similarities.

@@

\section{Acoustic Modeling}
\label{sec:am}
In this section, we summarize the different forms of acoustic model we use: 
the continuous, semi-continuous, multiple-codebook semi-continuous, and 
subspace forms of Gaussian Mixture Models. 
We also describe the phonetic decision-tree building process, which is 
necessary background for how we build the multiple-codebook systems.

\subsection{Acoustic-Phonetic Decision Tree Building}
The acoustic-phonetic decision tree provides a mapping from
a phonetic context window (e.g. a triphone), together with a HMM-state index, 
to an emission probability density function (pdf). 
%
The statistics for the tree clustering consist of the 
sufficient statistics needed to estimate a Gaussian for each seen combination 
of (phonetic context window, HMM-state).  These statistics are based on a Viterbi 
alignment of a previously built system.  The roots
of the decision trees can be configured; for the RM experiments, these correspond to
the phones in the system, and on Wall Street Journal, they
correspond to the ``real phones'', i.e. grouping different stress and
position-marked versions of the same phone together.

The splitting procedure can ask questions not only about the context phones,
but also about the central phone and the HMM state; the questions about phones are 
derived from an automatic clustering procedure.  The splitting procedure is
greedy and optimizes the likelihood of the data given a single Gaussian in each
tree leaf; this is subject to a fixed variance floor to avoid problems caused
by singleton counts and numerical roundoff.  We typically split up to a 
specified number of leaves and then cluster the resulting leaves as long as the
likelihood loss due to combining them is less than the likelihood improvement
on the final split of the tree-building process; and we restrict this 
clustering to disallow combining leaves across different subtrees (e.g., from 
different monophones).  The post-clustering process typically reduces the number 
of leaves by around 20\%.

@@

We initialize the mixtures with a single component each, and subsequently 
allocate more components by splitting components at every iteration until a 
pre-determined total number of components is reached. We allocate 
the Gaussians proportional to a small power (e.g.~0.2) of the state's 
occupation count.

@@

For better performance, we initialize the codebook using the tree statistics
collected on a prior phone alignment. For each tree leaf, we include
the respective Gaussian as a component in the mixture.
The components are merged and (if necessary) split to eliminate low count 
Gaussians and to match the desired number of components. In a final step, 
5 EM iterations are used to improve the goodness of fit to the acoustic 
training data.

\subsection{Multiple-Codebook Semi-continuous Models}
A style of system that lies between the semi-continuous and continuous types of
systems is one based on a two-level tree; this is described in~\cite{prasad2004t2b}
as a state-clustered tied-mixture (SCTM) system. The way we implement this is
to first build the decision tree used for phonetic clustering to a relatively
coarse level (e.g.~100 leaves). Each leaf of this tree corresponds to a codebook of Gaussians, i.e.~the Gaussian parameters are tied at this level. The leaves 
of this tree are then split further to give a more fine-grained clustering (e.g.~2500 leaves), where for each new leaf we remember which codebook it corresponds to. The leaves of this second tree correspond to the actual context-dependent HMM states and contain the weights.
For this type of system we do not apply any post-clustering.

@@

Similar to the traditional semi-continuous codebook initialization, we start
from the tree statistics and distribute the Gaussians to the respective codebooks
using the tree mapping to obtain preliminary codebooks.
The target size $N_k$ of codebook $k$ is determined with respect to 
the occupancies of the respective leaves as
\begin{equation}
N_k = N_0 + \frac
%% power rule, maybe some other day...
%  { \left(  \sum_{l \in \{m(l) = k\}} \text{occ}(l)  \right)^q }
%  { \sum_r \left(  \sum_{t \in \{m(t) = r\}} \text{occ}(l)  \right)^q } 
  { \sum_{l \in \{m(l) = k\}} \text{occ}(l) }
  { \sum_r \sum_{t \in \{m(t) = r\}} \text{occ}(l) } 
  \left( N - K \cdot N_0 \right)
\end{equation}
where $N_0$ is a minimum number of Gaussians per codebooks (e.g., 3), $N$ is
the total number of Gaussians, $K$ the number of codebooks, and $\text{occ}(j)$ 
is the occupancy of tree leaf $j$. 
%
%% Korbinian: No power rule for now, seems to harm results.
%, and $q$ controls the influence of the occupancy regarding one codebook. 
%A typical value of $q=0.2$ leads to rather homogeneous codebook sizes while 
%still attributing more Gaussians to states with larger occupancies. 
%For the two-level architecture, a $q=1$ yielded best results.
%
The target sizes of the codebooks are again enforced by either splitting or
merging the components.

\subsection{Subspace Gaussian Mixture Models}
The idea of subspace Gaussian mixture models (SGMM) is, similar to 
semi-continuous models, to reduce the number of parameters by selecting the
Gaussians from a subspace spanned by a universal background model (UBM)
and state specific transformations.   The SGMM emission pdfs can
be computed as
\begin{eqnarray}
p(\x | j) & = & \sum_{i=1}^{N} c_{ji} \nv(\x; \m_{ji}, \k_i) \\
\m_{ji}    & = & \M_i \v_j \\
c_{ji}     & = & \frac{\exp \w_i^T \v_j}{\sum_l^N \exp \w_l^T \v_j}
\end{eqnarray}
where the covariance matrices $\k_i$ are shared between all leaves $j$. The
weights $w_{ji}$ and means $\m_{ji}$ are derived from $\v_j$ together with
$\M_i$ and $\w_i$. The term ``subspace'' indicates that the parameters of the
mixtures are limited to a subspace of the entire space of parameters of
the underlying codebook.
%
A detailed description and derivation of the accumulation and update formulas
can be found in \cite{povey2011sgm}. Note that the experiments in this article
are without adaptation.

@@

The results on the different test sets on the RM data are displayed in 
Table~\ref{tab:res_rm}. The classic semi-continuous system shows the worst 
performance of the four, which serves to confirm the conventional wisdom
about the superiority of continous systems.
The two-level tree based multiple-codebook system performance lies in between the 
continuous and SGMM system.

@@

Keeping the number of codebooks and leaves as well as the smoothing parameters
$\tau$ and $\rho$ of {\em 2lvl} constant, we experiment with the number of
Gaussians and type of covariance; the results are displayed in 
Table~\ref{tab:rm_diagfull}.
Interestingly, the full covariances make a strong difference: Using the same
number of Gaussians, full covariances lead to a significant improvement.
On the other hand, a substantial increase of the number of diagonal Gaussians 
leads to a similar performance as the regular continuous system.
%
The effect of the two smoothing methods using $\tau$ and $\rho$ is not
clear from Table~\ref{tab:rm_diagfull}, i.e. the results are not always
consistent.  While the inter-iteration smoothing with $\rho$
helps in most cases, the intra-iteration smoothing coefficient $\tau$ 
shows an inconsistent effect.  In the future we may abandon this type of
smoothing or investigate other smoothing methods.

@@

The results on the different test sets on the WSJ data are displayed in
Table~\ref{tab:res_rm}.
% TODO Korbinian: The following is for the si-84/half results, as currently
%                 in the table.
Using the parameters calibrated on the RM test data, the {\em 2lvl} system
performance is not as good as the continuous or GMM systems, but still
in a similar range while using about the same number of leaves but about a
third of the Gaussians.
Once the number of Gaussians and leaves, as well as the smoothing parameters 
are tuned on the development set, we expect the error rates to be in between 
the continuous and SGMM systems, as seen on the RM data.

>>>>

\begin{abstract}
In the past decade, semi-continuous hidden Markov models (SC-HMMs) have not 
attracted much attention in the speech recognition community. Growing amounts 
of training data and increasing sophistication of model estimation led to the 
impression that continuous HMMs are the best choice of acoustic model.
%
However, recent work on recognition of under-resourced languages faces the same
old problem of estimating a large number of parameters from limited amounts 
of transcribed speech.
This has led to a renewed interest in methods of reducing the number of parameters 
while maintaining or extending the modeling capabilities of continuous models.
%
In this work, we compare classic and multiple-codebook semi-continuous models
using diagonal and full covariance matrices with continuous HMMs and subspace Gaussian 
mixture models.  
Experiments on the RM and WSJ corpora show that while a classical semi-continuous
system does not perform as well as a continuous one, multiple-codebook semi-continuous
systems can perform better, particular when using full-covariance Gaussians.
\end{abstract}

@@

We were motivated by a number of factors to look again at semi-continuous models.
One is that we are interested in applications where there is a small amount
of training data (or a small amount of in-domain training data), and since
in semi-continuous models the amount of state-specific parameters is relatively
small, we felt that they might have an advantage there.  We were also interested
in improving the open-source speech recognition toolkit {\sc Kaldi}~\cite{povey2011tks},
and felt that since a multiple-codebook form of semi-continuous models is, to
our knowledge, currently being used in a state-of-the-art system~\cite{prasad2004t2b},
we should implement this type of model.  We also wanted to experiment with
the style of semi-continuous system described 
in \cite{schukattalamazzini1994srf,schukattalamazzini1995as}, particularly the smoothing
techniques, and see how these compare against a more conventional system.
Lastly, we felt that it would be interesting to compare semi-continuous
models with Subspace Gaussian Mixture Models (SGMMs)~\cite{povey2011sgm}, since
there are obvious similarities.

@@

\section{Acoustic Modeling}
\label{sec:am}
In this section, we summarize the different forms of acoustic model we use: 
the continuous, semi-continuous, multiple-codebook semi-continuous, and 
subspace forms of Gaussian mixture models. 
We also describe the phonetic decision tree building process, which is 
necessary background for how we build the multiple-codebook systems.

\subsection{Acoustic-Phonetic Decision Tree Building}
The acoustic-phonetic decision tree provides a mapping from
a phonetic context window (e.g. a triphone), together with a HMM-state index, 
to an emission probability density function (pdf). 
%
The statistics for the tree clustering consist of the 
sufficient statistics needed to estimate a Gaussian for each seen combination 
of (phonetic context window, HMM-state).  These statistics are based on a Viterbi 
alignment of a previously built system.  The roots
of the decision trees can be configured; for the RM experiments, these correspond to
the phones in the system, and on Wall Street Journal, they
correspond to the ``real phones'', i.e., grouping different stress and
position-marked versions of the same phone together.

The splitting procedure can ask questions not only about the context phones,
but also about the central phone and the HMM state; the questions about phones are 
derived from an automatic clustering procedure.  The splitting procedure is
greedy and optimizes the likelihood of the data given a single Gaussian in each
tree leaf; this is subject to a fixed variance floor to avoid problems caused
by singleton counts and numerical roundoff.  We typically split up to a 
specified number of leaves and then cluster the resulting leaves as long as the
likelihood loss due to combining them is less than the likelihood improvement
on the final split of the tree building process; and we restrict this 
clustering to disallow combining leaves across different subtrees (e.g., from 
different monophones).  The post-clustering process typically reduces the number 
of leaves by around 20\%.

@@

We initialize the mixtures with a single component each, and subsequently 
allocate more components by splitting components at every iteration until a 
pre-determined total number of components is reached. We allocate 
the Gaussians proportional to a small power (e.g.,~0.2) of the state's 
occupation count.

@@

For better performance, we initialize the codebook using the tree statistics
collected on a prior Viterbi alignment. For each tree leaf, we include
the respective Gaussian as a component in the mixture.
The components are merged and (if necessary) split to eliminate low count 
Gaussians and to match the desired number of components. In a final step, 
5 EM iterations are used to improve the goodness of fit to the acoustic 
training data.

\subsection{Multiple-Codebook Semi-continuous Models}
A style of system that lies between the semi-continuous and continuous types of
systems is one based on a two-level tree; this is described in~\cite{prasad2004t2b}
as a state-clustered tied-mixture (SCTM) system. The way we implement this is
to first build the decision tree used for phonetic clustering to a relatively
coarse level (e.g.~100 leaves). Each leaf of this tree corresponds to a codebook 
of Gaussians, i.e.,~the Gaussian parameters are tied at this level. The leaves 
of this tree are then split further to give a more fine-grained clustering 
(e.g.,~2500 leaves), where for each new leaf we remember which codebook it corresponds to. The leaves of this second tree correspond to the actual context-dependent HMM states and contain the weights.
For this type of system we do not apply any post-clustering.

@@

Similar to the traditional semi-continuous codebook initialization, we start
from the tree statistics and distribute the Gaussians to the respective codebooks
using the tree mapping to obtain preliminary codebooks.
The target size $N_k$ of codebook $k$ is determined with respect to 
the occupancies of the respective leaves as
\begin{equation}
N_k = N_0 + \frac
%% power rule, maybe some other day...
%  { \left(  \sum_{l \in \{m(l) = k\}} \text{occ}(l)  \right)^q }
%  { \sum_r \left(  \sum_{t \in \{m(t) = r\}} \text{occ}(l)  \right)^q } 
  { \sum_{l \in \{m(l) = k\}} \text{occ}(l) }
  { \sum_t \text{occ}(t) } 
  \left( N - K \cdot N_0 \right)
\end{equation}
where $N_0$ is a minimum number of Gaussians per codebooks (e.g., 3), $N$ is
the total number of Gaussians, $K$ the number of codebooks, and $\text{occ}(j)$ 
is the occupancy of tree leaf $j$. 
%
%% Korbinian: No power rule for now, seems to harm results.
%, and $q$ controls the influence of the occupancy regarding one codebook. 
%A typical value of $q=0.2$ leads to rather homogeneous codebook sizes while 
%still attributing more Gaussians to states with larger occupancies. 
%For the two-level architecture, a $q=1$ yielded best results.
%
The target sizes of the codebooks are again enforced by either splitting or
merging the components.

\subsection{Subspace Gaussian Mixture Models}
The idea of subspace Gaussian mixture models (SGMM) is, similar to 
semi-continuous models, to reduce the number of parameters by selecting the
Gaussians from a subspace spanned by a universal background model (UBM)
and state specific transformations.   The SGMM emission pdfs can
be computed as
\begin{eqnarray}
p(\x | j) & = & \sum_{i=1}^{N} c_{ji} \nv(\x; \m_{ji}, \k_i) \\
\m_{ji}    & = & \M_i \v_j \\
c_{ji}     & = & \frac{\exp \w_i^T \v_j}{\sum_l^N \exp \w_l^T \v_j}
\end{eqnarray}
where the covariance matrices $\k_i$ are shared between all states $j$. The
weights $w_{ji}$ and means $\m_{ji}$ are derived from $\v_j$ together with
$\M_i$ and $\w_i$. The term ``subspace'' indicates that the parameters of the
mixtures are limited to a subspace of the entire space of parameters of
the underlying codebook.
%
A detailed description and derivation of the accumulation and update formulas
can be found in \cite{povey2011sgm}. Note that the experiments in this article
are without adaptation.

@@

The results on the different test sets on the RM data are displayed in 
Table~\ref{tab:res_rm}. The classic semi-continuous system shows the worst 
performance of the four, which serves to confirm the conventional wisdom
about the superiority of continuous systems.
The two-level tree based multiple-codebook system performance lies in between the 
continuous and SGMM system.

@@

Keeping the number of codebooks and leaves as well as the smoothing parameters
$\tau$ and $\rho$ of {\em 2lvl} constant, we experiment with the number of
Gaussians and type of covariance; the results are displayed in 
Table~\ref{tab:rm_diagfull}.
Interestingly, the full covariances make a strong difference: Using the same
number of Gaussians, full covariances lead to a significant improvement.
On the other hand, a substantial increase of the number of diagonal Gaussians 
leads to a similar performance as the regular continuous system.
%
The effect of the two smoothing methods using $\tau$ and $\rho$ is not
clear from Table~\ref{tab:rm_diagfull}, i.e., the results are not always
consistent.  While the inter-iteration smoothing with $\rho$
helps in most cases, the intra-iteration smoothing coefficient $\tau$ 
shows an inconsistent effect.  In the future we may abandon this type of
smoothing or investigate other smoothing methods.

@@

The results on the different test sets on the WSJ data are displayed in
Table~\ref{tab:res_rm}.
% TODO Korbinian: The following is for the si-84/half results, as currently
%                 in the table.
Using the parameters calibrated on the RM test data, the {\em 2lvl} system
performance is not as good as the continuous or SGMM systems, but still
in a similar range while using about the same number of leaves but about a
third of the Gaussians.
Once the number of Gaussians and leaves, as well as the smoothing parameters 
are tuned on the development set, we expect the error rates to be in between 
the continuous and SGMM systems, as seen on the RM data.
